---
title: "How to Clean a Messy Health Dataset"
subtitle: "A practical beginner’s guide to turning a messy spreadsheet into research-ready data using R."
author: "RISE Data Science"
date: "2026-06-24"
categories: [R, data cleaning, medical research, beginner]
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    code-fold: show
    code-summary: "Show code"
    code-tools: true
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  eval: false
  warning: false
  message: false
editor: visual
---

::: callout-note
## How to use this article

This article is a hands-on tutorial. It uses a fictional messy health dataset and walks through common data cleaning steps in R. The dataset is deliberately imperfect so that you can practise identifying and fixing realistic problems.

The dataset is fictional and created for teaching only. It should not be used for clinical decision-making.
:::

## Why data cleaning matters

In an ideal world, research datasets would arrive perfectly organised: clear column names, no missing values, no duplicates, no impossible entries, and all categories labelled consistently.

In reality, many datasets look more like a group assignment file at 2 am: useful, but slightly chaotic.

This is normal. The problem is not having messy data. The problem is analysing messy data without noticing.

Poor data cleaning can lead to:

- incorrect summary statistics;
- misleading p-values;
- biased results;
- duplicate participants being counted twice;
- impossible values being treated as real measurements;
- unclear methods that cannot be reproduced later.

For example, if missing blood pressure is coded as `999`, R may treat `999` as a real blood pressure unless you tell it otherwise. That would obviously distort the analysis.

Data cleaning is not just admin work. It is part of scientific reasoning. Before asking “Is this statistically significant?”, we need to ask “What does this dataset actually contain?”

## The messy sample dataset

This article uses a fictional dataset called `rise_sample_health_data_messy.csv`.

It contains participant-level health data with variables such as:

- participant ID;
- age;
- sex;
- BMI;
- systolic blood pressure;
- diastolic blood pressure;
- HbA1c in mmol/mol;
- smoking status;
- exercise minutes per week;
- diabetes status.

The dataset has intentionally messy problems, including:

- inconsistent sex labels such as `M`, `Male`, `male`, `F`, `Female`, and `female`;
- inconsistent smoking labels such as `Never`, `Non-smoker`, `Ex`, `Former`, and `Current`;
- missing values recorded as blank cells, `NA`, `Unknown`, `Not recorded`, `999`, and `-99`;
- impossible values such as negative age or extremely high blood pressure;
- duplicate participant IDs;
- messy column names with spaces and symbols.

The goal is not to make the dataset clinically perfect. The goal is to learn a safe, reproducible workflow for cleaning data.

## Preserve the raw dataset

Before touching the data, make one rule clear:

::: callout-important
## Never overwrite the raw dataset

Keep the original dataset unchanged. Create a cleaned version separately, and document every cleaning step in code.
:::

A sensible project structure is:

```{text}
rise_data_cleaning_tutorial/
├── data/
│   ├── rise_sample_health_data_messy.csv
│   └── rise_sample_health_data_cleaned.csv
├── scripts/
│   └── 01_clean_data.R
└── article_02_clean_messy_health_dataset.qmd
```

In real research, this separation matters. If a supervisor asks how a value was changed, or if a reviewer questions the cleaning decisions, you need to be able to trace the process.

A raw dataset is like the crime scene. Do not move the furniture before taking photos.

## Load packages and import the dataset

We will use two main packages:

- `tidyverse` for data manipulation and visualisation;
- `janitor` for cleaning column names and creating quick tables.

Install the packages if needed:

```{r}
install.packages("tidyverse")
install.packages("janitor")
```

Load the packages:

```{r}
library(tidyverse)
library(janitor)
```

Now import the messy dataset:

```{r}
health_raw <- read_csv("data/rise_sample_health_data_messy.csv")
```

At this point, `health_raw` should be treated as the imported raw version. We will inspect it before making changes.

## Inspect the dataset first

Before cleaning anything, inspect the dataset. This is where you look for suspicious patterns.

```{r}
head(health_raw)
glimpse(health_raw)
nrow(health_raw)
ncol(health_raw)
names(health_raw)
summary(health_raw)
```

Things to ask:

- How many rows are there?
- How many columns are there?
- What does each row represent?
- Are the column names easy to use in code?
- Are numeric variables being read as numbers or text?
- Are missing values coded consistently?
- Do any values look clinically impossible?
- Are there repeated participant IDs?

At this stage, write down what looks suspicious. Do not fix everything immediately. First, map the problems.

A useful first checklist:

- `Sex` appears to have inconsistent labels.
- `Smoking Status` appears to have inconsistent labels.
- Missing data appear to be recorded in several ways.
- Some numerical values look impossible.
- Column names contain spaces and symbols.
- Some participant IDs may be duplicated.

## Clean column names

Messy column names make code harder to write. Names such as `Participant ID`, `Systolic BP`, and `HbA1c (mmol/mol)` contain spaces, capitals, and symbols.

The `clean_names()` function from the `janitor` package converts names into a consistent format.

```{r}
health_clean <- health_raw %>%
  clean_names()

names(health_clean)
```

Example before and after:

| Before             | After             |
|--------------------|-------------------|
| `Participant ID`   | `participant_id`  |
| `Systolic BP`      | `systolic_bp`     |
| `HbA1c (mmol/mol)` | `hb_a1c_mmol_mol` |
| `Smoking Status`   | `smoking_status`  |

This may look cosmetic, but it prevents many annoying errors later.

## Handle missing values

Missing values are not always recorded as proper `NA`. In real datasets, missing values may appear as:

- blank cells;
- `NA`;
- `Unknown`;
- `Not recorded`;
- `999`;
- `-99`.

If these are not handled properly, R may treat them as real values or text labels.

One approach is to define the known missing-value codes when importing the dataset:

```{r}
missing_codes <- c("", "NA", "Unknown", "Not recorded", "999", "-99")

health_raw <- read_csv(
  "data/rise_sample_health_data_messy.csv",
  na = missing_codes
)

health_clean <- health_raw %>%
  clean_names()
```

Now count missing values in each column:

```{r}
health_clean %>%
  summarise(across(everything(), ~ sum(is.na(.))))
```

If the dataset has already been imported, you can also convert common missing-value labels after import:

```{r}
health_clean <- health_clean %>%
  mutate(
    across(
      where(is.character),
      ~ na_if(.x, "Unknown")
    ),
    across(
      where(is.character),
      ~ na_if(.x, "Not recorded")
    )
  )
```

Do not automatically delete all rows with missing data. First ask:

- Which variable is missing?
- How much data are missing?
- Why might the data be missing?
- Does the missingness affect the main research question?
- Should this decision be discussed with a supervisor or statistician?

For a beginner article, the key principle is simple: make missing values explicit and consistent.

## Check and convert variable types

Sometimes numerical variables are imported as text because they contain words such as `Unknown` or `Not recorded`.

After handling missing values, check the structure:

```{r}
glimpse(health_clean)
```

If needed, convert numerical columns:

```{r}
health_clean <- health_clean %>%
  mutate(
    age = as.numeric(age),
    bmi = as.numeric(bmi),
    systolic_bp = as.numeric(systolic_bp),
    diastolic_bp = as.numeric(diastolic_bp),
    hb_a1c_mmol_mol = as.numeric(hb_a1c_mmol_mol),
    exercise_minutes_per_week = as.numeric(exercise_minutes_per_week)
  )
```

If R creates warnings during conversion, inspect them. Warnings are not decorations; they are R tapping you on the shoulder.

## Standardise categorical variables

Categorical variables use labels rather than continuous numbers. Examples include sex, smoking status, and diabetes status.

The problem is that the same category may be written in several ways.

For example, sex might be recorded as:

- `M`;
- `Male`;
- `male`;
- `F`;
- `Female`;
- `female`.

We can standardise these into `Male` and `Female`.

```{r}
health_clean <- health_clean %>%
  mutate(
    sex = case_when(
      str_to_lower(sex) %in% c("m", "male") ~ "Male",
      str_to_lower(sex) %in% c("f", "female") ~ "Female",
      TRUE ~ NA_character_
    )
  )
```

Always check the result:

```{r}
tabyl(health_clean$sex)
```

Now standardise smoking status. We will collapse the messy labels into three groups: `Never`, `Former`, and `Current`.

```{r}
health_clean <- health_clean %>%
  mutate(
    smoking_status = str_to_lower(smoking_status),
    smoking_status = str_replace_all(smoking_status, "-", " "),
    smoking_status = case_when(
      smoking_status %in% c("never", "never smoker", "non smoker", "nonsmoker") ~ "Never",
      smoking_status %in% c("ex", "former", "former smoker") ~ "Former",
      smoking_status %in% c("current", "current smoker") ~ "Current",
      TRUE ~ NA_character_
    )
  )
```

Check the output:

```{r}
tabyl(health_clean$smoking_status)
```

The important habit is not just recoding. It is recoding and then checking.

## Check numerical ranges

Numerical variables may contain impossible or implausible values.

For this dataset, we should check:

- age;
- BMI;
- systolic blood pressure;
- diastolic blood pressure;
- HbA1c;
- exercise minutes per week.

Start with a range summary:

```{r}
health_clean %>%
  summarise(
    min_age = min(age, na.rm = TRUE),
    max_age = max(age, na.rm = TRUE),
    min_bmi = min(bmi, na.rm = TRUE),
    max_bmi = max(bmi, na.rm = TRUE),
    min_systolic_bp = min(systolic_bp, na.rm = TRUE),
    max_systolic_bp = max(systolic_bp, na.rm = TRUE),
    min_diastolic_bp = min(diastolic_bp, na.rm = TRUE),
    max_diastolic_bp = max(diastolic_bp, na.rm = TRUE),
    min_hba1c = min(hb_a1c_mmol_mol, na.rm = TRUE),
    max_hba1c = max(hb_a1c_mmol_mol, na.rm = TRUE)
  )
```

Then create flags for values needing review:

```{r}
health_clean <- health_clean %>%
  mutate(
    age_flag = !is.na(age) & (age < 0 | age > 120),
    bmi_flag = !is.na(bmi) & (bmi < 10 | bmi > 80),
    systolic_bp_flag = !is.na(systolic_bp) & (systolic_bp < 60 | systolic_bp > 250),
    diastolic_bp_flag = !is.na(diastolic_bp) & (diastolic_bp < 30 | diastolic_bp > 150),
    hba1c_flag = !is.na(hb_a1c_mmol_mol) & (hb_a1c_mmol_mol < 20 | hb_a1c_mmol_mol > 130)
  )
```

View the flagged rows:

```{r}
health_clean %>%
  filter(age_flag | bmi_flag | systolic_bp_flag | diastolic_bp_flag | hba1c_flag) %>%
  select(
    participant_id,
    age,
    bmi,
    systolic_bp,
    diastolic_bp,
    hb_a1c_mmol_mol,
    ends_with("flag")
  )
```

A flag does not automatically mean “delete this row.” It means “this needs review.” Some extreme values are real. Others are data entry errors. The decision depends on clinical context, the study protocol, and discussion with the research team.

## Check for duplicate participants

Duplicate rows can distort an analysis because some participants may be counted more than once.

Check for repeated participant IDs:

```{r}
duplicate_ids <- health_clean %>%
  count(participant_id) %>%
  filter(n > 1)

duplicate_ids
```

View the duplicated rows:

```{r}
health_clean %>%
  semi_join(duplicate_ids, by = "participant_id") %>%
  arrange(participant_id)
```

Do not delete duplicates automatically. There are several possible explanations:

- accidental duplicate entry;
- repeated measurement at different visits;
- baseline and follow-up rows;
- data merging error;
- two samples linked to the same participant.

The correct action depends on what each row is supposed to represent.

## Create useful derived variables

A derived variable is a new variable created from existing variables.

For example, we can create a BMI category:

```{r}
health_clean <- health_clean %>%
  mutate(
    bmi_category = case_when(
      is.na(bmi) ~ NA_character_,
      bmi < 18.5 ~ "Underweight",
      bmi >= 18.5 & bmi < 25 ~ "Healthy weight",
      bmi >= 25 & bmi < 30 ~ "Overweight",
      bmi >= 30 ~ "Obesity"
    )
  )
```

We can also create an example HbA1c group. The cut-offs below are included as a teaching example only; in real research, use the definitions specified by your study protocol, supervisor, or clinical guideline.

```{r}
health_clean <- health_clean %>%
  mutate(
    hba1c_group = case_when(
      is.na(hb_a1c_mmol_mol) ~ NA_character_,
      hb_a1c_mmol_mol < 42 ~ "Below diabetes threshold",
      hb_a1c_mmol_mol >= 42 & hb_a1c_mmol_mol < 48 ~ "Increased risk range",
      hb_a1c_mmol_mol >= 48 ~ "Diabetes range"
    )
  )
```

Check the derived variables:

```{r}
tabyl(health_clean$bmi_category)
tabyl(health_clean$hba1c_group)
```

Derived variables should always be documented. Different cut-offs can change the results, so these decisions should not be hidden.

## Create a simple cleaning log

A cleaning log records what was changed and why. This is useful for supervisors, collaborators, methods sections, and your own sanity.

A simple cleaning log might look like this:

| Step | Problem found | Action taken | Reason |
|------------------|------------------|------------------|------------------|
| 1 | Column names had spaces and symbols | Used `clean_names()` | Easier and safer coding |
| 2 | Missing values had several labels | Converted known missing codes to `NA` | Consistent missing-value handling |
| 3 | Sex labels inconsistent | Re-coded to `Male` and `Female` | Standardised categories |
| 4 | Smoking labels inconsistent | Re-coded to `Never`, `Former`, and `Current` | Standardised categories |
| 5 | Implausible numerical values found | Created flag variables | Values need review before exclusion |
| 6 | Duplicate participant IDs found | Flagged duplicated IDs | Need to determine whether true duplicates or repeated measures |

You can keep this log in a spreadsheet, a Word document, comments inside your R script, or a section of your Quarto report.

## Save the cleaned dataset

Once the cleaning steps are complete, save the cleaned dataset separately from the raw dataset.

```{r}
write_csv(health_clean, "data/rise_sample_health_data_cleaned.csv")
```

Also save your R script, for example:

```{text}
scripts/01_clean_data.R
```

The aim is that someone else could start from the raw dataset, run your script, and reproduce the cleaned dataset.

## Create a simple summary table

After cleaning, you can begin producing simple descriptive summaries. For example, summarise age and BMI by diabetes status:

```{r}
health_clean %>%
  group_by(diabetes_status) %>%
  summarise(
    n = n(),
    mean_age = mean(age, na.rm = TRUE),
    median_age = median(age, na.rm = TRUE),
    mean_bmi = mean(bmi, na.rm = TRUE),
    median_bmi = median(bmi, na.rm = TRUE)
  )
```

This is not yet a full statistical analysis. It is a descriptive table that helps you understand the dataset after cleaning.

You can also create a basic plot:

```{r}
ggplot(health_clean, aes(x = diabetes_status, y = bmi)) +
  geom_boxplot() +
  labs(
    title = "BMI by diabetes status",
    x = "Diabetes status",
    y = "BMI"
  )
```

## Final data cleaning checklist

Before moving to statistical analysis, check that:

- the raw data have been preserved;
- the cleaned data are saved separately;
- column names are clean and consistent;
- missing values are consistently coded as `NA`;
- categorical variables are standardised;
- numerical variables have been checked for impossible values;
- suspicious values are flagged rather than blindly deleted;
- duplicate participant IDs have been checked;
- derived variables are documented;
- the cleaning script has been saved;
- uncertain decisions have been discussed with a supervisor or statistician.

## Mini activity: clean the RISE sample dataset

Download `rise_sample_health_data_messy.csv` and work through the following tasks.

### Task

1.  Load the dataset into R.
2.  Clean the column names.
3.  Identify missing values.
4.  Standardise the sex variable.
5.  Standardise smoking status.
6.  Convert numerical variables into the correct type if needed.
7.  Flag impossible age, BMI, blood pressure, and HbA1c values.
8.  Check for duplicate participant IDs.
9.  Create a BMI category.
10. Save the cleaned dataset.

### Starter code

```{r}
library(tidyverse)
library(janitor)

missing_codes <- c("", "NA", "Unknown", "Not recorded", "999", "-99")

health_raw <- read_csv(
  "data/rise_sample_health_data_messy.csv",
  na = missing_codes
)

health_clean <- health_raw %>%
  clean_names()

health_clean <- health_clean %>%
  mutate(
    sex = case_when(
      str_to_lower(sex) %in% c("m", "male") ~ "Male",
      str_to_lower(sex) %in% c("f", "female") ~ "Female",
      TRUE ~ NA_character_
    ),
    smoking_status = str_to_lower(smoking_status),
    smoking_status = str_replace_all(smoking_status, "-", " "),
    smoking_status = case_when(
      smoking_status %in% c("never", "never smoker", "non smoker", "nonsmoker") ~ "Never",
      smoking_status %in% c("ex", "former", "former smoker") ~ "Former",
      smoking_status %in% c("current", "current smoker") ~ "Current",
      TRUE ~ NA_character_
    ),
    age_flag = !is.na(age) & (age < 0 | age > 120),
    bmi_flag = !is.na(bmi) & (bmi < 10 | bmi > 80),
    systolic_bp_flag = !is.na(systolic_bp) & (systolic_bp < 60 | systolic_bp > 250),
    diastolic_bp_flag = !is.na(diastolic_bp) & (diastolic_bp < 30 | diastolic_bp > 150),
    hba1c_flag = !is.na(hb_a1c_mmol_mol) & (hb_a1c_mmol_mol < 20 | hb_a1c_mmol_mol > 130),
    bmi_category = case_when(
      is.na(bmi) ~ NA_character_,
      bmi < 18.5 ~ "Underweight",
      bmi >= 18.5 & bmi < 25 ~ "Healthy weight",
      bmi >= 25 & bmi < 30 ~ "Overweight",
      bmi >= 30 ~ "Obesity"
    )
  )

health_clean %>% summarise(across(everything(), ~ sum(is.na(.))))
tabyl(health_clean$sex)
tabyl(health_clean$smoking_status)

health_clean %>%
  count(participant_id) %>%
  filter(n > 1)

write_csv(health_clean, "data/rise_sample_health_data_cleaned.csv")
```

### Reflection questions

- Which data problems were obvious immediately?
- Which problems were easier to find using R than Excel?
- Which values would you flag but not delete immediately?
- Which decisions would need supervisor input?
- How would you describe your cleaning process in a methods section?

## What comes next?

Once the dataset is clean, we can start asking research questions. Should we compare two groups? Test an association? Build a regression model? Create a figure for a poster?

A clean dataset does not guarantee a good analysis, but a messy dataset can absolutely ruin one.

::: callout-tip
## Key takeaway

Data cleaning is where research analysis becomes trustworthy. Preserve the raw data, document every step, and treat suspicious values as questions to investigate rather than problems to erase.
:::

## Suggested downloadable resources

- [`rise_sample_health_data_messy.csv`](data/rise_sample_health_data_messy.csv)
- [`rise_sample_health_data_cleaned.csv`](data/rise_sample_health_data_cleaned.csv)
- Example cleaning script
- Data cleaning checklist
- Cleaning log template
