---
title: "Descriptive Statistics and Table 1 in R"
subtitle: "How to summarise cleaned clinical research data before formal analysis"
author: "RISE Education Team"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    code-fold: show
    code-summary: "Show code"
    code-tools: true
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  eval: true
  warning: false
  message: false
editor: visual
---

::: callout-note
## How to use this article

This article builds on Articles 1 and 2.

Article 1 introduced R basics. Article 2 cleaned a messy clinical research dataset. In this article, we use the cleaned dataset to create descriptive summaries, a simple Table 1, exploratory plots, and first group comparisons.
:::

## 1. Why descriptive statistics come before statistical tests

Before running formal statistical tests, we need to understand the dataset.

Descriptive statistics help us answer basic questions:

- How many participants are in the dataset?
- What is the typical age, BMI, blood pressure, or HbA1c?
- How many participants have diabetes?
- Are there missing values?
- Are the groups roughly similar or clearly different?
- Do the distributions look sensible?

This is not just a box-ticking exercise. Descriptive statistics are how we notice problems before they become embarrassing. Better to find a strange value now than after making a poster.

## 2. Load packages

We will use the same beginner-friendly packages as before.

```{r}
#| eval: false
# Run this only if the packages are not already installed.
install.packages("tidyverse")
install.packages("janitor")
```

```{r}
library(tidyverse)
library(janitor)

dir.create("outputs", showWarnings = FALSE)
dir.create("figures", showWarnings = FALSE)
```

## 3. Import the cleaned dataset

This article uses the cleaned dataset from Article 2.

```{r}
data_path <- "data/rise_sample_health_data_cleaned.csv"

if (!file.exists(data_path)) {
  stop("Cannot find the cleaned dataset. Check that rise_sample_health_data_cleaned.csv is inside the data/ folder.")
}

clean_data <- read_csv(data_path, show_col_types = FALSE)
```

Inspect the data.

```{r}
glimpse(clean_data)
head(clean_data)
```

## 4. Decide which variables to summarise

For this article, we will separate variables into two broad groups.

```{r}
continuous_vars <- c(
  "age",
  "bmi",
  "systolic_bp",
  "diastolic_bp",
  "hba1c_mmol_mol",
  "exercise_minutes_per_week"
)

categorical_vars <- c(
  "sex",
  "smoking_status",
  "diabetes_status",
  "recruitment_site",
  "study_group"
)
```

Continuous variables are numeric measurements. Categorical variables are groups or labels.

## 5. Count missing values

Even after cleaning, some missing values may remain. That is normal. The important thing is to know where they are.

```{r}
missing_summary <- clean_data %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_n"
  ) %>%
  arrange(desc(missing_n))

missing_summary
```

```{r}
write_csv(missing_summary, "outputs/article_03_missing_summary.csv")
```

## 6. Summarise continuous variables

For continuous variables, common summaries include:

- number of non-missing observations;
- mean;
- standard deviation;
- median;
- interquartile range;
- minimum and maximum.

```{r}
continuous_summary <- clean_data %>%
  summarise(
    across(
      all_of(continuous_vars),
      list(
        n = ~ sum(!is.na(.x)),
        mean = ~ mean(.x, na.rm = TRUE),
        sd = ~ sd(.x, na.rm = TRUE),
        median = ~ median(.x, na.rm = TRUE),
        q1 = ~ quantile(.x, 0.25, na.rm = TRUE),
        q3 = ~ quantile(.x, 0.75, na.rm = TRUE),
        min = ~ min(.x, na.rm = TRUE),
        max = ~ max(.x, na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("variable", ".value"),
    names_pattern = "(.*)_(n|mean|sd|median|q1|q3|min|max)$"
  ) %>%
  mutate(
    across(where(is.numeric), ~ round(.x, 1))
  )

continuous_summary
```

```{r}
write_csv(continuous_summary, "outputs/article_03_continuous_summary.csv")
```

## 7. Summarise categorical variables

For categorical variables, we usually report counts and percentages.

```{r}
categorical_summary <- clean_data %>%
  select(all_of(categorical_vars)) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "category"
  ) %>%
  filter(!is.na(category)) %>%
  count(variable, category, name = "n") %>%
  group_by(variable) %>%
  mutate(
    percent = round(100 * n / sum(n), 1),
    result = paste0(n, " (", percent, "%)")
  ) %>%
  ungroup()

categorical_summary
```

```{r}
write_csv(categorical_summary, "outputs/article_03_categorical_summary.csv")
```

## 8. What is a Table 1?

A **Table 1** is a baseline characteristics table. It is common in clinical papers.

It usually describes the study population using variables such as age, sex, BMI, comorbidities, blood pressure, and laboratory values.

A Table 1 is not mainly about proving a hypothesis. It helps readers understand who is in the study.

## 9. Create a simple overall Table 1

To keep this article beginner-friendly, we will create a simple Table 1 using `dplyr` rather than specialised packages.

First, create a formatted table for continuous variables.

```{r}
table1_continuous <- continuous_summary %>%
  transmute(
    variable,
    summary = paste0(
      median,
      " [",
      q1,
      ", ",
      q3,
      "]"
    ),
    format = "Median [IQR]"
  )

table1_continuous
```

Now create a formatted table for categorical variables.

```{r}
table1_categorical <- categorical_summary %>%
  transmute(
    variable = paste0(variable, ": ", category),
    summary = result,
    format = "n (%)"
  )

table1_categorical
```

Combine both into one table.

```{r}
table1_overall <- bind_rows(
  table1_continuous,
  table1_categorical
)

table1_overall
```

```{r}
write_csv(table1_overall, "outputs/article_03_table1_overall.csv")
```

## 10. Create Table 1 by diabetes status

Often, we want to describe the dataset by a key grouping variable. Here, we will use `diabetes_status`.

First, summarise continuous variables by diabetes status.

```{r}
table1_by_diabetes_continuous <- clean_data %>%
  group_by(diabetes_status) %>%
  summarise(
    across(
      all_of(continuous_vars),
      list(
        median = ~ median(.x, na.rm = TRUE),
        q1 = ~ quantile(.x, 0.25, na.rm = TRUE),
        q3 = ~ quantile(.x, 0.75, na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = -diabetes_status,
    names_to = c("variable", ".value"),
    names_pattern = "(.*)_(median|q1|q3)$"
  ) %>%
  mutate(
    summary = paste0(
      round(median, 1),
      " [",
      round(q1, 1),
      ", ",
      round(q3, 1),
      "]"
    )
  ) %>%
  select(variable, diabetes_status, summary) %>%
  pivot_wider(names_from = diabetes_status, values_from = summary)

table1_by_diabetes_continuous
```

Now summarise categorical variables by diabetes status.

```{r}
table1_by_diabetes_categorical <- clean_data %>%
  select(diabetes_status, all_of(categorical_vars)) %>%
  select(-diabetes_status, everything(), diabetes_status) %>%
  pivot_longer(
    cols = all_of(setdiff(categorical_vars, "diabetes_status")),
    names_to = "variable",
    values_to = "category"
  ) %>%
  filter(!is.na(category), !is.na(diabetes_status)) %>%
  count(diabetes_status, variable, category, name = "n") %>%
  group_by(diabetes_status, variable) %>%
  mutate(
    percent = round(100 * n / sum(n), 1),
    summary = paste0(n, " (", percent, "%)")
  ) %>%
  ungroup() %>%
  mutate(variable = paste0(variable, ": ", category)) %>%
  select(variable, diabetes_status, summary) %>%
  pivot_wider(names_from = diabetes_status, values_from = summary)

table1_by_diabetes_categorical
```

Combine them.

```{r}
table1_by_diabetes <- bind_rows(
  table1_by_diabetes_continuous,
  table1_by_diabetes_categorical
)

table1_by_diabetes
```

```{r}
write_csv(table1_by_diabetes, "outputs/article_03_table1_by_diabetes_status.csv")
```

## 11. Visualise key variables

Descriptive tables are useful, but plots often reveal patterns more quickly.

### BMI distribution

```{r}
bmi_plot <- ggplot(clean_data, aes(x = bmi)) +
  geom_histogram(binwidth = 2, colour = "white") +
  labs(
    title = "Distribution of BMI",
    x = "BMI",
    y = "Number of participants"
  ) +
  theme_minimal()

bmi_plot
```

```{r}
ggsave("figures/article_03_bmi_distribution.png", bmi_plot, width = 7, height = 5, dpi = 300)
```

### HbA1c by diabetes status

```{r}
hba1c_plot <- ggplot(clean_data, aes(x = diabetes_status, y = hba1c_mmol_mol)) +
  geom_boxplot() +
  geom_jitter(width = 0.15, alpha = 0.6) +
  labs(
    title = "HbA1c by diabetes status",
    x = "Diabetes status",
    y = "HbA1c (mmol/mol)"
  ) +
  theme_minimal()

hba1c_plot
```

```{r}
ggsave("figures/article_03_hba1c_by_diabetes_status.png", hba1c_plot, width = 7, height = 5, dpi = 300)
```

### BMI and HbA1c

```{r}
bmi_hba1c_plot <- ggplot(clean_data, aes(x = bmi, y = hba1c_mmol_mol)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Relationship between BMI and HbA1c",
    x = "BMI",
    y = "HbA1c (mmol/mol)"
  ) +
  theme_minimal()

bmi_hba1c_plot
```

```{r}
ggsave("figures/article_03_bmi_hba1c_scatter.png", bmi_hba1c_plot, width = 7, height = 5, dpi = 300)
```

## 12. First simple group comparisons

Descriptive statistics tell us what the data look like. A statistical test asks whether an observed difference is larger than we might expect by chance, under a particular set of assumptions.

In this beginner article, the goal is not to memorise tests. The goal is to understand the logic.

::: callout-warning
## Do not rush this step

A statistical test should follow the research question, study design, variable type, and assumptions. Do not choose tests by habit alone.
:::

## 13. Compare BMI by diabetes status

BMI is numeric. Diabetes status has two groups: `Yes` and `No`.

A simple first comparison is a two-sample t-test.

```{r}
bmi_t_test <- t.test(bmi ~ diabetes_status, data = clean_data)

bmi_t_test
```

We can store the key result in a small table.

```{r}
bmi_t_test_summary <- tibble(
  comparison = "BMI by diabetes status",
  test = "Two-sample t-test",
  p_value = bmi_t_test$p.value
)

bmi_t_test_summary
```

```{r}
write_csv(bmi_t_test_summary, "outputs/article_03_bmi_t_test_summary.csv")
```

## 14. Compare sex distribution by diabetes status

Sex and diabetes status are both categorical variables. A simple first comparison is a chi-square test.

```{r}
sex_diabetes_table <- table(clean_data$sex, clean_data$diabetes_status)

sex_diabetes_table
```

```{r}
sex_diabetes_chisq <- chisq.test(sex_diabetes_table)

sex_diabetes_chisq
```

Save the result.

```{r}
sex_diabetes_chisq_summary <- tibble(
  comparison = "Sex by diabetes status",
  test = "Chi-square test",
  p_value = sex_diabetes_chisq$p.value
)

sex_diabetes_chisq_summary
```

```{r}
write_csv(sex_diabetes_chisq_summary, "outputs/article_03_sex_diabetes_chisq_summary.csv")
```

::: callout-note
## A note about p-values

A p-value does not tell us whether a result is clinically important. It only helps assess how compatible the observed data are with a null hypothesis under the assumptions of the test.

Clinical judgement, study design, sample size, effect size, confidence intervals, and prior evidence all matter.
:::

## 15. Add confidence intervals for a mean

Confidence intervals are often more informative than p-values alone. Here is a simple example using mean BMI.

```{r}
bmi_mean_ci <- clean_data %>%
  summarise(
    n = sum(!is.na(bmi)),
    mean_bmi = mean(bmi, na.rm = TRUE),
    sd_bmi = sd(bmi, na.rm = TRUE),
    standard_error = sd_bmi / sqrt(n),
    lower_95_ci = mean_bmi - 1.96 * standard_error,
    upper_95_ci = mean_bmi + 1.96 * standard_error
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

bmi_mean_ci
```

```{r}
write_csv(bmi_mean_ci, "outputs/article_03_bmi_mean_confidence_interval.csv")
```

## 16. Export a simple analysis summary

It is useful to save important outputs rather than relying only on what appears in the console.

```{r}
analysis_summary <- bind_rows(
  bmi_t_test_summary,
  sex_diabetes_chisq_summary
)

analysis_summary
```

```{r}
write_csv(analysis_summary, "outputs/article_03_simple_analysis_summary.csv")
```

## 17. Mini activity

Using the cleaned dataset, try to answer these questions.

1.  What is the median age of the sample?
2.  What percentage of participants have diabetes?
3.  What is the median HbA1c by diabetes status?
4.  Make a boxplot of systolic blood pressure by diabetes status.
5.  Create a simple Table 1 by study group.
6.  Decide whether each variable is continuous, categorical, binary, or ordinal.
7.  Write one sentence describing the study population.

## 18. Common beginner mistakes

| Mistake | Why it matters |
|----|----|
| Running tests before looking at the data | You may miss obvious errors or strange distributions |
| Reporting only p-values | P-values do not show size or clinical importance |
| Ignoring missing values | Different variables may have different sample sizes |
| Using the wrong denominator for percentages | Percentages can become misleading |
| Treating Table 1 as an afterthought | Table 1 is often the first thing readers use to understand your study |

## 19. What comes next?

You now have a clean beginner workflow:

1.  Article 1: learn R basics and inspect a clean dataset.
2.  Article 2: clean a messy clinical dataset reproducibly.
3.  Article 3: summarise the cleaned dataset and create first tables, plots, and comparisons.

A natural next article would focus on choosing statistical tests more carefully, including:

- continuous vs categorical outcomes;
- paired vs unpaired data;
- parametric vs non-parametric tests;
- confidence intervals;
- simple linear and logistic regression.
