---
title: "Cleaning Clinical Research Data in R"
subtitle: "How to turn a messy spreadsheet into an analysis-ready dataset"
author: "RISE Education Team"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    code-fold: show
    code-summary: "Show code"
    code-tools: true
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  eval: true
  warning: false
  message: false
editor: visual
---

::: callout-note
## How to use this article

This article builds on Article 1. It assumes that you can open RStudio, run code chunks, load packages, import a CSV file, inspect a dataset, and read simple pipe-based code.

The goal here is to learn how to clean a messy research dataset in a reproducible way.
:::

## 1. From clean teaching data to messy real-world data

In Article 1, we used a clean fictional health dataset to learn the basics of R. Real clinical research datasets are rarely that tidy.

A spreadsheet may look organised at first glance, but it can still contain problems that quietly damage the analysis.

Common problems include:

- inconsistent column names;
- missing values;
- duplicate participants;
- inconsistent category labels;
- impossible biological values;
- numbers accidentally stored as text;
- unclear variable coding.

Before asking "Which statistical test should I run?", we first need to ask: **Can I trust this dataset?**

## 2. The core rule: never overwrite the raw dataset

The raw dataset is the original record. Do not overwrite it.

A better workflow is:

1.  keep the raw dataset untouched;
2.  write cleaning steps in R code;
3.  save a separate cleaned dataset;
4.  keep a short record of important cleaning decisions.

::: callout-tip
## Main idea

The raw dataset should stay unchanged. Clean the data using code and save a new cleaned version.
:::

## 3. The messy sample dataset

In this article, we will use a fictional teaching dataset called `rise_sample_health_data_messy.csv`.

It is a messy version of the clean dataset from Article 1. The problems are deliberate, because this is how we learn.

| Variable | What it represents | Example problem |
|------------------------|------------------------|------------------------|
| `Participant ID` | Participant identifier | Extra spaces and duplicate IDs |
| `Age` | Age in years | Missing values and impossible values |
| `Sex` | Sex recorded in the dataset | `Male`, `male`, `M`, `F`, unclear values |
| `BMI` | Body mass index | Missing values and impossible values |
| `Systolic BP` | Systolic blood pressure | Values such as `142 mmHg` |
| `HbA1c (mmol/mol)` | HbA1c | Values such as `48 mmol/mol` |
| `Smoking Status` | Smoking history | `Never`, `N`, `ex-smoker`, inconsistent labels |
| `Diabetes Status` | Diabetes status | `Yes`, `Y`, `T2DM`, `No`, inconsistent labels |

The aim is to turn this messy raw dataset into an analysis-ready dataset.

## 4. Project folder structure for data cleaning

For data cleaning, we need a slightly more careful folder structure than in Article 1.

``` text
article_02_data_cleaning_in_r/
├── article_02_cleaning_clinical_research_data.qmd
├── data_raw/
│   └── rise_sample_health_data_messy.csv
├── data_cleaned/
│   └── rise_sample_health_data_cleaned.csv
├── scripts/
│   └── article_02_cleaning_code.R
└── outputs/
    ├── data_cleaning_summary.csv
    ├── duplicate_participant_report.csv
    ├── missing_value_summary.csv
    ├── suspicious_values_report.csv
    └── variable_dictionary.csv
```

The key idea is simple:

- `data_raw/` contains the original dataset;
- `data_cleaned/` contains the cleaned dataset;
- `scripts/` contains reusable R code;
- `outputs/` contains reports generated during cleaning.

## 5. Load packages and create folders

We will use two main packages.

- `tidyverse` helps us import, clean, summarise, and visualise data.
- `janitor` helps us clean column names and create quick tables.

```{r}
#| eval: false
# Run this only if the packages are not already installed.
install.packages("tidyverse")
install.packages("janitor")
```

```{r}
# Load packages for this session.
library(tidyverse)
library(janitor)

# Create output folders if they do not already exist.
dir.create("data_cleaned", showWarnings = FALSE)
dir.create("outputs", showWarnings = FALSE)
```

## 6. Import the raw dataset safely

In a messy dataset, it is often useful to import everything as text first. This prevents R from guessing variable types too early.

```{r}
raw_data_path <- "data_raw/rise_sample_health_data_messy.csv"

if (!file.exists(raw_data_path)) {
  stop("Cannot find the raw dataset. Check that it is inside the data_raw/ folder.")
}

raw_data <- read_csv(
  file = raw_data_path,
  col_types = cols(.default = col_character()),
  na = c("", "NA", "N/A", "Unknown", "Not recorded", "not available")
)
```

At this stage, `raw_data` should be treated as the untouched raw dataset.

## 7. Inspect the raw dataset

Article 1 introduced basic inspection commands. Here, we use them with a different purpose: to find cleaning problems.

```{r}
head(raw_data)
```

```{r}
glimpse(raw_data)
```

```{r}
dim(raw_data)
names(raw_data)
```

Do not clean anything yet. First, look carefully.

## 8. Count missing values before cleaning

It is useful to count missing values before making changes. This gives us a baseline record.

```{r}
missing_before_cleaning <- raw_data %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_n"
  ) %>%
  arrange(desc(missing_n))

missing_before_cleaning
```

This code creates a table showing how many missing values are present in each variable.

## 9. Clean column names

Raw clinical spreadsheets often have column names with spaces, capital letters, brackets, or symbols.

R can work with those names, but they are awkward to type. We can use `clean_names()` from the `janitor` package to make them easier to use.

```{r}
clean_data <- raw_data %>%
  clean_names()

names(clean_data)
```

For example:

``` text
Participant ID        -> participant_id
Systolic BP           -> systolic_bp
HbA1c (mmol/mol)      -> hb_a1c_mmol_mol
```

For consistency in this teaching series, we will rename `hb_a1c_mmol_mol` to `hba1c_mmol_mol`.

```{r}
clean_data <- clean_data %>%
  rename_with(
    .fn = ~ str_replace(.x, "^hb_a1c_mmol_mol$", "hba1c_mmol_mol")
  )

names(clean_data)
```

## 10. Clean participant IDs and extra spaces

Extra spaces can create hidden problems. For example, `RISE001` and `RISE001` may look similar to a human, but R treats them as different values.

```{r}
clean_data <- clean_data %>%
  mutate(
    participant_id = str_squish(participant_id),
    participant_id = str_to_upper(participant_id),
    across(where(is.character), str_squish)
  )
```

This step:

- removes repeated spaces;
- removes spaces at the start and end of text;
- standardises participant IDs to uppercase.

## 11. Check duplicate participants

A duplicate participant ID means the same participant appears more than once.

```{r}
duplicate_participants <- clean_data %>%
  count(participant_id, name = "n_rows") %>%
  filter(n_rows > 1)

duplicate_participants
```

Save this report before removing duplicates.

```{r}
write_csv(duplicate_participants, "outputs/duplicate_participant_report.csv")
```

For this teaching dataset, we will keep the first row for each participant.

```{r}
n_before_duplicates <- nrow(clean_data)

clean_data <- clean_data %>%
  distinct(participant_id, .keep_all = TRUE)

n_after_duplicates <- nrow(clean_data)
```

::: callout-warning
## Be careful in real research

In real clinical data, duplicate IDs are not always errors. They may represent repeated visits, multiple admissions, or longitudinal follow-up. Always check the study design before removing duplicates.
:::

## 12. Convert numeric variables

Some values look numeric but are actually stored as text. For example, `142 mmHg` is text, not a number.

The function `parse_number()` extracts the numeric part from a text value.

```{r}
clean_data <- clean_data %>%
  mutate(
    age = parse_number(age),
    bmi = parse_number(bmi),
    systolic_bp = parse_number(systolic_bp),
    diastolic_bp = parse_number(diastolic_bp),
    hba1c_mmol_mol = parse_number(hba1c_mmol_mol),
    exercise_minutes_per_week = parse_number(exercise_minutes_per_week)
  )
```

Numeric variables are needed for means, medians, plots, and statistical tests.

## 13. Standardise categorical variables

Categorical variables often contain inconsistent labels.

For example, the same idea may be recorded as `Male`, `male`, `M`, or `man`.

We can standardise these using `str_to_lower()` and `case_when()`.

```{r}
clean_data <- clean_data %>%
  mutate(
    sex = str_to_lower(sex),
    sex = case_when(
      sex %in% c("male", "m", "man") ~ "Male",
      sex %in% c("female", "f", "woman") ~ "Female",
      TRUE ~ NA_character_
    ),
    diabetes_status = str_to_lower(diabetes_status),
    diabetes_status = case_when(
      diabetes_status %in% c("yes", "y", "t2dm", "type 2 diabetes", "diabetes") ~ "Yes",
      diabetes_status %in% c("no", "n", "no diabetes") ~ "No",
      TRUE ~ NA_character_
    ),
    smoking_status = str_to_lower(smoking_status),
    smoking_status = case_when(
      smoking_status %in% c("never", "n", "never smoker") ~ "Never",
      smoking_status %in% c("former", "ex-smoker", "past smoker") ~ "Former",
      smoking_status %in% c("current", "c", "smoker") ~ "Current",
      TRUE ~ NA_character_
    ),
    recruitment_site = str_to_title(recruitment_site),
    study_group = str_to_title(study_group)
  )
```

Check the cleaned categories.

```{r}
clean_data %>% count(sex)
clean_data %>% count(diabetes_status)
clean_data %>% count(smoking_status)
```

## 14. Check impossible or suspicious values

This is where medical judgement matters. Some values may be technically numeric but biologically implausible.

For this teaching dataset, we will use simple example cut-offs.

```{r}
suspicious_values <- bind_rows(
  clean_data %>%
    filter(age < 0 | age > 120) %>%
    transmute(participant_id, variable = "age", value = as.character(age), reason = "Age outside 0-120"),
  clean_data %>%
    filter(bmi < 10 | bmi > 80) %>%
    transmute(participant_id, variable = "bmi", value = as.character(bmi), reason = "BMI outside 10-80"),
  clean_data %>%
    filter(systolic_bp < 70 | systolic_bp > 250) %>%
    transmute(participant_id, variable = "systolic_bp", value = as.character(systolic_bp), reason = "Systolic BP outside 70-250"),
  clean_data %>%
    filter(diastolic_bp < 40 | diastolic_bp > 150) %>%
    transmute(participant_id, variable = "diastolic_bp", value = as.character(diastolic_bp), reason = "Diastolic BP outside 40-150"),
  clean_data %>%
    filter(hba1c_mmol_mol < 15 | hba1c_mmol_mol > 160) %>%
    transmute(participant_id, variable = "hba1c_mmol_mol", value = as.character(hba1c_mmol_mol), reason = "HbA1c outside 15-160"),
  clean_data %>%
    filter(exercise_minutes_per_week < 0 | exercise_minutes_per_week > 2000) %>%
    transmute(participant_id, variable = "exercise_minutes_per_week", value = as.character(exercise_minutes_per_week), reason = "Exercise minutes outside 0-2000")
)

suspicious_values
```

```{r}
write_csv(suspicious_values, "outputs/suspicious_values_report.csv")
```

::: callout-important
## Teaching example only

These cut-offs are used for teaching. In real research, plausibility rules should be discussed with supervisors, clinicians, statisticians, or the study protocol.
:::

## 15. Replace impossible values with missing values

After creating a report, we can replace impossible values with `NA`.

```{r}
clean_data <- clean_data %>%
  mutate(
    age = if_else(age < 0 | age > 120, NA_real_, age),
    bmi = if_else(bmi < 10 | bmi > 80, NA_real_, bmi),
    systolic_bp = if_else(systolic_bp < 70 | systolic_bp > 250, NA_real_, systolic_bp),
    diastolic_bp = if_else(diastolic_bp < 40 | diastolic_bp > 150, NA_real_, diastolic_bp),
    hba1c_mmol_mol = if_else(hba1c_mmol_mol < 15 | hba1c_mmol_mol > 160, NA_real_, hba1c_mmol_mol),
    exercise_minutes_per_week = if_else(
      exercise_minutes_per_week < 0 | exercise_minutes_per_week > 2000,
      NA_real_,
      exercise_minutes_per_week
    )
  )
```

We are not pretending those values never existed. We have already saved them in `suspicious_values_report.csv`.

## 16. Count missing values after cleaning

Now repeat the missing value count.

```{r}
missing_after_cleaning <- clean_data %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_n"
  ) %>%
  arrange(desc(missing_n))

missing_after_cleaning
```

```{r}
write_csv(missing_after_cleaning, "outputs/missing_value_summary.csv")
```

## 17. Create a cleaning summary

A cleaning summary helps readers understand what changed.

```{r}
data_cleaning_summary <- tibble(
  cleaning_step = c(
    "Raw dataset",
    "Before removing duplicate participant IDs",
    "After removing duplicate participant IDs",
    "Final cleaned dataset"
  ),
  n_rows = c(
    nrow(raw_data),
    n_before_duplicates,
    n_after_duplicates,
    nrow(clean_data)
  )
)

data_cleaning_summary
```

```{r}
write_csv(data_cleaning_summary, "outputs/data_cleaning_summary.csv")
```

## 18. Create a variable dictionary

A **variable dictionary** explains what each variable means. This is a very useful habit in research projects.

```{r}
variable_dictionary <- tribble(
  ~variable, ~description, ~type, ~cleaning_note,
  "participant_id", "Unique fictional participant identifier", "Identifier", "Cleaned to uppercase and stripped of extra spaces",
  "age", "Age in years", "Continuous numeric", "Values outside 0-120 replaced with missing",
  "sex", "Sex recorded in the dataset", "Categorical", "Standardised to Male/Female; unclear values set to missing",
  "bmi", "Body mass index in kg/m^2", "Continuous numeric", "Values outside 10-80 replaced with missing",
  "systolic_bp", "Systolic blood pressure in mmHg", "Continuous numeric", "Values outside 70-250 replaced with missing",
  "diastolic_bp", "Diastolic blood pressure in mmHg", "Continuous numeric", "Values outside 40-150 replaced with missing",
  "hba1c_mmol_mol", "HbA1c in mmol/mol", "Continuous numeric", "Values outside 15-160 replaced with missing",
  "smoking_status", "Smoking history", "Categorical", "Standardised to Never/Former/Current",
  "diabetes_status", "Diabetes status", "Binary categorical", "Standardised to Yes/No",
  "recruitment_site", "Fictional recruitment location", "Categorical", "Converted to title case",
  "study_group", "Fictional study group", "Categorical", "Converted to title case",
  "exercise_minutes_per_week", "Self-reported exercise minutes per week", "Continuous numeric", "Values outside 0-2000 replaced with missing"
)

variable_dictionary
```

```{r}
write_csv(variable_dictionary, "outputs/variable_dictionary.csv")
```

## 19. Save the cleaned dataset

Now save the cleaned dataset separately.

```{r}
write_csv(clean_data, "data_cleaned/rise_sample_health_data_cleaned.csv")
```

The raw dataset remains in `data_raw/`. The cleaned dataset is now in `data_cleaned/`.

## 20. Final check of the cleaned data

Always inspect the cleaned dataset before moving on.

```{r}
glimpse(clean_data)
summary(clean_data)
```

```{r}
clean_data %>% tabyl(sex)
clean_data %>% tabyl(smoking_status)
clean_data %>% tabyl(diabetes_status)
```

A simple plot can also help check whether the cleaned data looks sensible.

```{r}
ggplot(clean_data, aes(x = bmi)) +
  geom_histogram(binwidth = 2, colour = "white") +
  labs(
    title = "Distribution of BMI after cleaning",
    x = "BMI",
    y = "Number of participants"
  ) +
  theme_minimal()
```

## 21. Mini activity

Using the messy dataset, try to reproduce the cleaning workflow yourself.

1.  Import the raw dataset.
2.  Inspect the first few rows.
3.  Count missing values.
4.  Clean column names.
5.  Clean participant IDs.
6.  Find duplicate participant IDs.
7.  Convert numeric variables.
8.  Standardise categorical variables.
9.  Check impossible or suspicious values.
10. Save the cleaned dataset.

## 22. Common beginner mistakes

| Mistake | Why it matters |
|-----------------------------------|-----------------------------------|
| Overwriting the raw dataset | You lose the original record |
| Deleting all missing values too early | You may remove too much data without thinking |
| Assuming duplicate IDs are always errors | They may represent repeated visits or follow-up |
| Cleaning manually in Excel without a record | The work becomes hard to check or repeat |
| Changing values without documenting why | Later readers cannot understand your decisions |

## 23. What comes next?

Now that the dataset is clean, we can start asking research questions.

In Article 3, we will use the cleaned dataset to create descriptive summaries, build a simple Table 1, compare groups, and make cleaner exploratory figures.
